---
title: "Deliverable 1"
author: "Carlson Smith"
date: "10/15/2020"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(ggplot2)
```

# Introduction

In this project I will be studying fires, electronic, climate change, and more
with their effects on carbon emissions on earth.  Using these predictions or 
data I will try to look to the future as well and see if those correlations can 
be used to predict temperatures.

The Domains I will specifically be looking into are:

* Fire Emissions
* Fire frequency
* Temperature History
* Carbon Emissions
* Clean Energy
* Electrical Usage

# Data Set

```{r}
co2.data <- read_csv("owid-co2-data.xlsx.csv")
head(co2.data)
```
This data comes from https://github.com/owid/co2-data that is maintained by Our World in Data. It is updated regularly and includes data on CO2 emissions along with other helpful metrics.

The Columns for this cut down data set are:

* iso_code - 	ISO 3166-1 alpha-3 – three-letter country codes
* country- Geographic location
* year - Year of observation
* co2 - Annual production-based emissions of carbon dioxide (CO2), measured in million tonnes per year.
* co2_growth_prct - Percentage change in CO2 emissions from one year relative to the previous year.
* co2_growth_abs - 	Annual change in CO2 emissions from one year relative to the previous year, measured in million tonnes.
* share_global_co2 - National or regional annual CO2 emissions, measured as a percentage of the global total
* cumulative_co2 - Cumulative emissions of CO2 from 1751 through to the given year, measured in million tonnes.
* share_global_cumulative_co2 - National or regional annual cumulative CO2 emissions, measured as a share of the global total
* cement_co2 - CO2 emissions from cement production, measured in million tonnes.
* coal_co2 - CO2 emissions from coal production, measured in million tonnes.
* flaring_co2 - CO2 emissions from gas flaring, measured in million tonnes.
* gas_co2	- CO2 emissions from gas production, measured in million tonnes.
* oil_co2 - CO2 emissions from oil production, measured in million tonnes.
* population - Total population
* gdp - Total real gross domestic product, inflation-adjusted


This graph shows the carbon emissions for countries based on the year.
```{r}
ggplot(data = co2.data, mapping = aes(x=year, y=co2))+
  geom_point(alpha = 0.2)
```
You can see an explosive growth in the 1970's with some countries climbing higher than others.
This data is just a base line to work with for future comparisons and predictions




















