---
title: "Deliverable 2"
author: "Carlson Smith"
date: "10/15/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(ggplot2)
options(warn=-1)
```

# Data from part 1

```{r}
co2.data <- read_csv("owid-co2-data.xlsx.csv")
tree.data <- read_csv("global_country_tree_cover_loss.csv")
fire.data <- read_csv("US_Fires.csv")
car.data <- read_csv("Annual_Emissions_per_Vehicle.csv")
electric.data <- read_csv("Electricity_Sources.csv")
co2.data <- filter(co2.data, iso_code != is.na(iso_code), co2 > 0)
tree.30threshold <- filter(tree.data, threshold == 30)
names(tree.30threshold)[7] <- "2001"
names(tree.30threshold)[8] <- "2002"
names(tree.30threshold)[9] <- "2003"
names(tree.30threshold)[10] <- "2004"
names(tree.30threshold)[11] <- "2005"
names(tree.30threshold)[12] <- "2006"
names(tree.30threshold)[13] <- "2007"
names(tree.30threshold)[14] <- "2008"
names(tree.30threshold)[15] <- "2009"
names(tree.30threshold)[16] <- "2010"
names(tree.30threshold)[17] <- "2011"
names(tree.30threshold)[18] <- "2012"
names(tree.30threshold)[19] <- "2013"
names(tree.30threshold)[20] <- "2014"
names(tree.30threshold)[21] <- "2015"
names(tree.30threshold)[22] <- "2016"
names(tree.30threshold)[23] <- "2017"
names(tree.30threshold)[24] <- "2018"
names(tree.30threshold)[25] <- "2019"
tree.pivot <- pivot_longer(tree.30threshold,c('2001','2002','2003','2004','2005','2006','2007','2008','2009','2010','2011','2012','2013','2014','2015','2016','2017','2018','2019'),names_to = "year", values_to = "tc_loss_ha")
head(tree.pivot)
tree.pivot$year <- as.double(tree.pivot$year)
tree.co2.data <- left_join(co2.data, tree.pivot) %>%
  filter(year > 2000)
```
Need to look for data of electric car sales in each country, could have relation
to co2 output depending on the power grid of a country.  If so I can 


# Data Science Questions

Looking at this initial data relating to each other with the whole world its co2 
data, to just looking at the US to find a closer relation, it seems like 
deforestation might be promising to explore to see if that is more closely related 
to emissions.

I also want to look into GPD and co2 output as well, i think that they don't have 
to be connected, but could also be in close relation when your economy is based 
off of machinery that isn't the cleanest.  For example a country that has a higher 
oil co2 release might have a higher GPD and be having more deforestation as well.
This would all point to a mindset though rather than an actual mathematical pattern 
that you can follow.

I don't think there are any ethical or social disadvantages to looking into this
area and digging deeper into relations between the forests of our world and climate 
change.  If anything, this research will end up pointing fingers and showing the 
countries that really need to get their act together and push society further along 
to improve itself.

# Possible Model Topics.

``` {r}
ggplot(data = filter(co2.data, country == "World"), mapping = aes(x=year, y=co2, color = country))+
  geom_smooth(alpha = 0.2)+
  ggtitle("World co2 output")
ggplot(data = filter(co2.data, country == "World"), mapping = aes(x=year, y=gdp, color = country))+
  geom_smooth(alpha = 0.2)+
  ggtitle("World overall GPD")
```
From the data science questions I wanted to look into from the initial data analysis, it seems like gdp and co2 might have a very close relation based on the general look of the world.  THough I do want to look into specific countries since I know the USA has curbed their co2 outputs but their gdp is still rising.  

THe model would be a simple relation between gdp and the co2 outputs
```{r}
ggplot(data = filter(co2.data, country == "United States"), mapping = aes(x=co2, y=gdp))+
  geom_point(alpha = 0.2)+
  ggtitle("GDP vs Co2")
```
While its quite hard to see much in this point graph, looking into individual co2 emissions might provide better results.
