---
title: "Deliverable 1"
author: "Carlson Smith"
date: "10/15/2020"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(ggplot2)
```

# Introduction

In this project I will be studying fires, electronic, climate change, and more with their effects on carbon emissions on earth.  Using these predictions or data I will try to look to the future as well as find other correlations between the climate and how we live.  The reason I'm interested in this field of research is because of the fires and reduction of snow in California.  My parents have talked about how much snow there used to be in Tahoe, and how the fires have just gotten bigger.  I want to know what we can change, and what causes the climate to warm and change over time

The Domains I will specifically be looking into are:

* Fire Emissions
* Fire frequency
* Temperature History
* Carbon Emissions
* Clean Energy
* Electrical Usage

# Data Sets

```{r}
co2.data <- read_csv("owid-co2-data.xlsx.csv")
```
co2.data comes from https://github.com/owid/co2-data that is maintained by Our World in Data. It is updated regularly and includes data on CO2 emissions along with other helpful metrics.  The original data set has a few more continuous variables that I didn't think I would need for what I want to look at.

The categorical variables for co2.data are:

* iso_code - 	ISO 3166-1 alpha-3 – three-letter country codes
* country - Geographic location
* year - Year of observation

The continuous variables for co2.data are:

* co2 - Annual production-based emissions of carbon dioxide (CO2), measured in million tonnes per year.
* co2_growth_prct - Percentage change in CO2 emissions from one year relative to the previous year.
* co2_growth_abs - 	Annual change in CO2 emissions from one year relative to the previous year, measured in million tonnes.
* share_global_co2 - National or regional annual CO2 emissions, measured as a percentage of the global total
* cumulative_co2 - Cumulative emissions of CO2 from 1751 through to the given year, measured in million tonnes.
* share_global_cumulative_co2 - National or regional annual cumulative CO2 emissions, measured as a share of the global total
* cement_co2 - CO2 emissions from cement production, measured in million tonnes.
* coal_co2 - CO2 emissions from coal production, measured in million tonnes.
* flaring_co2 - CO2 emissions from gas flaring, measured in million tonnes.
* gas_co2	- CO2 emissions from gas production, measured in million tonnes.
* oil_co2 - CO2 emissions from oil production, measured in million tonnes.
* population - Total population
* gdp - Total real gross domestic product, inflation-adjusted

co2.data comes from Our World in Data and will have some limitations, but should be a very trustworthy source as it does come from a source with a good reputation.  You also know that the website doesn't make money off of what they do since it's a .org site.  The one downside of this data set is that they don't split up co2 emissions to all the possible sources which means the countries actual co2 release could be different.

```{r}
tree.data <- read_csv("global_country_tree_cover_loss.csv")
```

tree.data comes from https://www.globalforestwatch.org that is made through a partnership with World Resources Institute. It is updated regularly and keeps track of the forests starting from around 2001.  Since it's hard to keep track of forests before the initial photos in 2001 were taken, the data is a little more limited to recent years

The categorical variables for tree.data are:

* country - Geographic location
* threshold - Percent canopy cover levels

The continuous variables for tree.data are:

* area_ha - Hectares of tree cover
* extent_2000_ha - Hectares of tree cover in 2000
* extent_2010_ha - Hectares of tree cover in 2010
* gain_2000-2012_ha - From 2001 to 2012, hectares of tree cover gained
* tc_loss_ha_2001 - Hectares of tree cover loss for 2001
* All years in between show loss for different years
* tc_loss_ha_2019 - Hectares tree cover loss for 2019

tree.data is another organization who monitors the forests of our world with satellite data.  Again, they are an .org so the website isn't made to get money and the data they get is going to be quite accurate. Using high resolution satellite imagery to look at our forests is a great way to get a generic summary of how they're doing without too much work. 

One worry for this data set is that comparison between 2001-2009 and 2011-2019 is not supposed to be very accurate or show patterns. The authors of the data said to preform this with caution making me think that they are comparing new data to extent 2000 or extent 2010 to find the loss of tree cover.  If comparing the two, tree growth and forest regrowth will have to be considered.

```{r}
car.data <- read_csv("Annual_Emissions_per_Vehicle.csv")
electric.data <- read_csv("Electricity_Sources.csv")
```

car.data and electric.data is grabbed from https://afdc.energy.gov/vehicles/electric_emissions.html that breaks down the data into a pie chart and bar chart.  This data source is to fill in some blanks of data in our previous data sets so we can better explore our questions

The categorical variables for car.data are:

* car_type - Style of engine for cars

The continuous variables for car.data are:

* pounds_of_c02 - pounds of co2 released by car over a year
* miles_driven - miles driven by car over a year

The categorical variables for electric.data are:

* electric_sources - Electric source for the grid

The continuous variables for electric.data are:

* percentage_grid - percentage of the grid supported by type of source

These two sets of data are from the United States Department of Energy who also listed where they got their estimations and data.  The sources they used and the results of the summarization all come from quite trustworthy sources.  Though the one downside of this data set is that it is only for the United States and not global.  

# Exploratory Data Analysis

This graph shows the carbon emissions for countries based on the year.
```{r}
head(co2.data)
summary(co2.data)
```
You can see an explosive growth in the 1970's with some countries climbing higher than others.
This data is just a base line to work with for future comparisons and predictions.

```{r}
ggplot(data = co2.data, mapping = aes(x=year, y=co2))+
  geom_point(alpha = 0.2)
  # Something here to add title to plot
```

```{r}
head(tree.data)
summary(tree.data)
```

tree.date looks at all of the tree cover, and to make it a little easier to find patterns in the data, I'm going to cut down the canopy density percentage to 30% and up.  Along with that, I'm going to rotate the data to make year a new variable so I can easily graph tree loss over the years.

```{r}
tree.30threshold <- filter(tree.data, threshold == 30)
names(tree.30threshold)[7] <- "2001"
names(tree.30threshold)[8] <- "2002"
names(tree.30threshold)[9] <- "2003"
names(tree.30threshold)[10] <- "2004"
names(tree.30threshold)[11] <- "2005"
names(tree.30threshold)[12] <- "2006"
names(tree.30threshold)[13] <- "2007"
names(tree.30threshold)[14] <- "2008"
names(tree.30threshold)[15] <- "2009"
names(tree.30threshold)[16] <- "2010"
names(tree.30threshold)[17] <- "2011"
names(tree.30threshold)[18] <- "2012"
names(tree.30threshold)[19] <- "2013"
names(tree.30threshold)[20] <- "2014"
names(tree.30threshold)[21] <- "2015"
names(tree.30threshold)[22] <- "2016"
names(tree.30threshold)[23] <- "2017"
names(tree.30threshold)[24] <- "2018"
names(tree.30threshold)[25] <- "2019"
tree.pivot <- pivot_longer(tree.30threshold,c('2001','2002','2003','2004','2005','2006','2007','2008','2009','2010','2011','2012','2013','2014','2015','2016','2017','2018','2019'),names_to = "years", values_to = "tc_loss_ha")
head(tree.pivot)
summary(tree.pivot)
```

```{r}
head(car.data)
head(electric.data)
```


















