---
title: "Deliverable 1"
author: "Carlson Smith"
date: "10/15/2020"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(ggplot2)
options(warn=-1)
```

# Introduction

In this project I will be studying climate change, deforestation, fires, and more
with their effects on carbon emissions on earth.  Using these predictions or data
I will try to look to the future as well as find other correlations between the 
climate and how we live.  The reason I'm interested in this field of research is 
because of the fires and reduction of snow in California along with the fear of 
how our actions might effect future generations.  My parents have talked about how
much snow there used to be in Tahoe, and how the fires have just gotten bigger.  
I want to know what we can change, and what causes the climate to warm and change
over time

The Domains I will specifically be looking into are:

* Fire Emissions
* Fire frequency
* Temperature History
* Carbon Emissions
* Clean Energy
* Electrical Usage

# Data Sets

```{r}
co2.data <- read_csv("owid-co2-data.xlsx.csv")
```
co2.data comes from https://github.com/owid/co2-data that is maintained by Our 
World in Data. It is updated regularly and includes data on CO2 emissions along 
with other helpful metrics.  The original data set has a few more continuous 
variables that I didn't think I would need for what I want to look at.

The categorical variables for co2.data are:

* iso_code - 	ISO 3166-1 alpha-3 – three-letter country codes
* country - Geographic location
* year - Year of observation

The continuous variables for co2.data are:

* co2 - Annual production-based emissions of carbon dioxide (CO2), measured in million tonnes per year.
* co2_growth_prct - Percentage change in CO2 emissions from one year relative to the previous year.
* co2_growth_abs - 	Annual change in CO2 emissions from one year relative to the previous year, measured in million tonnes.
* share_global_co2 - National or regional annual CO2 emissions, measured as a percentage of the global total
* cumulative_co2 - Cumulative emissions of CO2 from 1751 through to the given year, measured in million tonnes.
* share_global_cumulative_co2 - National or regional annual cumulative CO2 emissions, measured as a share of the global total
* cement_co2 - CO2 emissions from cement production, measured in million tonnes.
* coal_co2 - CO2 emissions from coal production, measured in million tonnes.
* flaring_co2 - CO2 emissions from gas flaring, measured in million tonnes.
* gas_co2	- CO2 emissions from gas production, measured in million tonnes.
* oil_co2 - CO2 emissions from oil production, measured in million tonnes.
* population - Total population
* gdp - Total real gross domestic product, inflation-adjusted

co2.data comes from Our World in Data and will have some limitations, but should
be a very trustworthy source as it does come from a source with a good reputation.
You also know that the website doesn't make money off of what they do since it's 
a .org site.  The one downside of this data set is that they don't split up co2 
emissions to all the possible sources which means the countries actual co2 release
could be different.

```{r}
tree.data <- read_csv("global_country_tree_cover_loss.csv")
```

tree.data comes from https://www.globalforestwatch.org that is made through a 
partnership with World Resources Institute. It is updated regularly and keeps 
track of the forests starting from around 2001.  Since it's hard to keep track 
of forests before the initial photos in 2001 were taken, the data is a little 
more limited to recent years.

The categorical variables for tree.data are:

* country - Geographic location
* threshold - Percent canopy cover levels

The continuous variables for tree.data are:

* area_ha - Hectares of tree cover
* extent_2000_ha - Hectares of tree cover in 2000
* extent_2010_ha - Hectares of tree cover in 2010
* gain_2000-2012_ha - From 2001 to 2012, hectares of tree cover gained
* tc_loss_ha_2001 - Hectares of tree cover loss for 2001
* All years in between show loss for different years
* tc_loss_ha_2019 - Hectares tree cover loss for 2019

tree.data is another organization who monitors the forests of our world with 
satellite data.  Again, they are an .org so the website isn't made to get money 
and the data they get is going to be quite accurate. Using high resolution 
satellite imagery to look at our forests is a great way to get a generic summary 
of how they're doing without too much work. 

One worry for this data set is that comparison between 2001-2009 and 2011-2019 is 
not supposed to be very accurate or show patterns. The authors of the data said to 
preform this with caution making me think that they are comparing new data to extent
2000 or extent 2010 to find the loss of tree cover.  If comparing the two, tree 
growth and forest regrowth will have to be considered.

```{r}
fire.data <- read_csv("US_Fires.csv")
```

fire.data comes from https://www.nifc.gov/fireInfo/fireInfo_stats_totalFires.html.
The data for this data set comes from the National Interagency Fire Center who 
handles multi agency corrdination for fire fighting in the United States.  The data
that it provides is a general summary of the US and the fires over years.

The categorical variables for fire.data are:

* Year - The year fire data relates to

The continuous variables for fire.data are:

* Fires - Number of fires
* Acres - Acres burned

fire.data does come from a .org and national agency, but I dont know how reliable
it is going to be.  I have never heard of this group until I was looking for data
so I might run into issues later if I realize this data isn't reliable.

Another worry with this data is there is a chance deforestation is the US and 
the acres burned are connected, but also who knows, more research is going to be
needed to figure out if they track the similar areas.

```{r}
car.data <- read_csv("Annual_Emissions_per_Vehicle.csv")
electric.data <- read_csv("Electricity_Sources.csv")
```

car.data and electric.data is grabbed from 
https://afdc.energy.gov/vehicles/electric_emissions.html that breaks down the 
data into a pie chart and bar chart.  This data source is to fill in some blanks 
of data in our previous data sets so we can better explore our questions

The categorical variables for car.data are:

* car_type - Style of engine for cars

The continuous variables for car.data are:

* pounds_of_c02 - pounds of co2 released by car over a year
* miles_driven - miles driven by car over a year

The categorical variables for electric.data are:

* electric_sources - Electric source for the grid

The continuous variables for electric.data are:

* percentage_grid - percentage of the grid supported by type of source

These two sets of data are from the United States Department of Energy who also 
listed where they got their estimations and data.  The sources they used and the 
results of the summarization all come from quite trustworthy sources.  Though the 
one downside of this data set is that it is only for the United States and not global.  

# Data Prep and Analysis

co2.data has data that covers generic land regions that overlap with countries.  
This is just added up from each country in it so I figured by deleting the data 
that didn't have an iso_code, we could remove some of this overlap.  I'm also 
going to delete any data that doesn't have co2 output since that's what we really \
are about when looking at this data
```{r}
co2.data <- filter(co2.data, iso_code != is.na(iso_code), co2 > 0)
head(co2.data)
```
This graph shows the carbon emissions for countries based on the year, though the
first time I tried this, it didn't work since there were too many countries with 
very low carbon output to draw lines for each one.  For ease of visualization, I'm 
going to set the co2 to above 100 million tonnes per year.
```{r}
ggplot(data = filter(co2.data, co2 > 100, country != "World"), mapping = aes(x=year, y=co2, color = country))+
  geom_smooth(alpha = 0.2)+
  ggtitle("co2 outputs of countries")
ggplot(data = filter(co2.data, country == "World"), mapping = aes(x=year, y=co2, color = country))+
  geom_smooth(alpha = 0.2)+
  ggtitle("World co2 output")
```

You can see from this data set, the overall growth of co2 emissions and outputs 
with the World summary line.  Admittedly there are too many countries to see all 
of this added up on the graph, but you can see the general idea with the visible 
lines.  There are also lots more in the data set that are not put in the graph 
since we wouldn't be able to graph each line, these are just the biggest emitters.
Though you can see each line with their growth over the years, some moveing faster
than others.

```{r}
ggplot()+
  geom_smooth(data = filter(co2.data, country == "World"), se = FALSE, mapping = aes(x=year, y=cement_co2, color='cement'))+
  geom_smooth(data = filter(co2.data, country == "World"), se = FALSE, mapping = aes(x=year, y=coal_co2, color='coal'))+
  geom_smooth(data = filter(co2.data, country == "World"), se = FALSE, mapping = aes(x=year, y=flaring_co2, color='flaring'))+
  geom_smooth(data = filter(co2.data, country == "World"), se = FALSE, mapping = aes(x=year, y=gas_co2, color='gas'))+
  geom_smooth(data = filter(co2.data, country == "World"), se = FALSE, mapping = aes(x=year, y=oil_co2, color='oil'))+
  ggtitle("World coal co2 output")
```

This graph splits up all the sources of co2 in the world to really show who is 
doing the most and what really needs to be fixed if we want to reduce our emissions.
These could all be things fixed in the future to reverse the damage we have done 
so far.

```{r}
head(tree.data)
```

tree.date looks at all of the tree cover, and to make it a little easier to find 
patterns in the data, I'm going to cut down the canopy density percentage to 30% 
and up.  The reason for cutting at 30% is that its the better visualization of tree 
loss, you can't choose 100%, but you cant choose 10%, as acres lost at that point 
would do very little to the number of trees.  Along with that, I'm going to rotate 
the data to make year a new variable so I can easily graph tree loss over the years 
as well as join co2 and trees later on.

```{r}
tree.30threshold <- filter(tree.data, threshold == 30)
names(tree.30threshold)[7] <- "2001"
names(tree.30threshold)[8] <- "2002"
names(tree.30threshold)[9] <- "2003"
names(tree.30threshold)[10] <- "2004"
names(tree.30threshold)[11] <- "2005"
names(tree.30threshold)[12] <- "2006"
names(tree.30threshold)[13] <- "2007"
names(tree.30threshold)[14] <- "2008"
names(tree.30threshold)[15] <- "2009"
names(tree.30threshold)[16] <- "2010"
names(tree.30threshold)[17] <- "2011"
names(tree.30threshold)[18] <- "2012"
names(tree.30threshold)[19] <- "2013"
names(tree.30threshold)[20] <- "2014"
names(tree.30threshold)[21] <- "2015"
names(tree.30threshold)[22] <- "2016"
names(tree.30threshold)[23] <- "2017"
names(tree.30threshold)[24] <- "2018"
names(tree.30threshold)[25] <- "2019"
tree.pivot <- pivot_longer(tree.30threshold,c('2001','2002','2003','2004','2005','2006','2007','2008','2009','2010','2011','2012','2013','2014','2015','2016','2017','2018','2019'),names_to = "year", values_to = "tc_loss_ha")
head(tree.pivot)
```

Using this pivot, we can combine with the co2 data with the years and countries 
that appear in both data sets.  This gives us a generic and overarching data set 
with deforestation rates and co2 emissions.  I'm going to filter as well since if 
I don't want to join and have tree data being empty, I want the related years only, 
which is limited my tree.pivot which goes 2001-2019.

```{r}
tree.pivot$year <- as.double(tree.pivot$year)
tree.co2.data <- left_join(co2.data, tree.pivot) %>%
  filter(year > 2000)
head(tree.co2.data)
```


To take a peek for relations in the tree/co2 data set, I'm going to reduce it to 
just the USA where I can look at both the growth rates of deforestation and how 
much co2 we make

```{r}
ggplot(data = filter(tree.co2.data, iso_code == 'USA'), mapping = aes(x=year, y=co2, color = country))+
  geom_smooth(alpha = 0.2)+
  ggtitle("CO2 released for the US over the years")
ggplot(data = filter(tree.co2.data, iso_code == 'USA'), mapping = aes(x=year, y=tc_loss_ha, color = country))+
  geom_smooth(alpha = 0.2)+
  ggtitle("Tree cover lost over the years")
ggplot(data = filter(tree.co2.data, iso_code == 'USA'), mapping = aes(x=year, y=gdp, color = country))+
  geom_smooth(alpha = 0.2)+
  ggtitle("GDP over the years")
```

while it does look like deforestation and co2 output seem to have very similar 
curves on this graph pair at first, it could be explored more to see if this is 
true with other countries as well.  If this is true then I would have to look into 
the idea of co2 output and hectares lost being connected on the co2 charts.



```{r}
head(fire.data)
summary(fire.data)
```

The Fire data for the US is mainly to see if there is a connection with the co2 
emissions and rates of fire as well.  There is a chance that deforestation reduces 
fires as well which could be something explored in the future as well.  This could 
be explored further if I look into the Australian data with their recent mega fire.

```{r}
ggplot()+
  geom_smooth(data = fire.data, se=FALSE, mapping = aes(x=Year, y=Fires, color = 'Fires'))+
  ggtitle("US Fires over years")
ggplot()+
  geom_smooth(data = fire.data, se=FALSE, mapping = aes(x=Year, y=Acres, color = 'Acres Burned'))+
  ggtitle("US Acres Burned")
```

Though this data is only the United States, it is interesting seeing the dip end 
in 1975, and slowly go back up while the number of fires still goes down, it seems 
like the fires are just getting bigger in more recent years, but we have less fires 
starting overall.

```{r}
head(car.data)
head(electric.data)
```

These last two data sets are just to fill in later when we look at renewable 
energy and what would happen if we managed to switch to a green house gas free 
life style which would include changing the power grid and switching to electric 
cars.  This data set needs to be worked on with more tables if I pursue this path 
further.

# Data Science Questions

Looking at this initial data relating to each other with the whole world its co2 
data, to just looking at the US to find a closer relation, it seems like 
deforestation might be promising to explore to see if that is more closely related 
to emissions.

I also want to look into GPD and co2 output as well, i think that they don't have 
to be connected, but could also be in close relation when your economy is based 
off of machinery that isn't the cleanest.  For example a country that has a higher 
oil co2 release might have a higher GPD and be having more deforestation as well.
This would all point to a mindset though rather than an actual mathematical pattern 
that you can follow.

I don't think there are any ethical or social disadvantages to looking into this
area and digging deeper into relations between the forests of our world and climate 
change.  If anything, this research will end up pointing fingers and showing the 
countries that really need to get their act together and push society further along 
to improve itself.















